A new simplified manual tour, with examples in mathematica and R

### 2022-07-06 paper/paper.tex
trying to map out the paper

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, can some variables be removed from the
projection without affecting the clustering. For interpreting models,
one can reduce or increase a variable's contribution to examine the
variable importance. These controls can also be used to interactively
generate facetted plots \citep{XXX}, or spatiotemporal glyphmaps
\citep{XXX}. Having the user interact with a projection is extremely
valuable for understanding high-dimensional data. However, these
algorithms have two problems: (1) the pre-processing of creating a
manipulation space overly complicates the algorithm, (2) extending to
higher dimensional control is difficult.

@@

\hypertarget{sec:method}{%
\section{Manual tour}\label{sec:method}}

An orthonormal basis (\(A_{p\times d}\)) and a variable id
(\(m \in \{1, ..., p\}\)) to control are provided to initialise a manual
tour. A method to update the values of the component of the controlled
variable \(V_m\) is needed.

\hypertarget{background}{%
\subsection{Background}\label{background}}

@@

\hypertarget{new-simpler-definition}{%
\subsection{New simpler definition}\label{new-simpler-definition}}

@@

The problem with Algorithm 1 is that the precise values for \(V_m\)
cannot be specified because the orthonormalisation wil change them. This
modification will maintain the components of \(V_m\) precisely (Figure
\ref{fig:manualsequence} (bottom row)). The algorithm is as follows:

@@

If \(d>2\) the process would be sequentially repeated in the same manner
that Gram-Schmidt is applied sequentially to orthormalise the columns of
a matrix. If \(d=1\) no orthonormalisation is needed, and the projection
vector would simply need to be normalised after each adjustment.

@@

\begin{enumerate}
\item Cursor movement takes ${x_{m1},x_{m2}}\to {a,b}$
\item The freedom to change the components $A_{i3}\cdots A_{ip}$ (the columns of $A$ not corresponding to the projection matrix) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $A_{m3}=\sqrt{1-a^2-b^2},~A_{m,k>3}=0$
\item For other rows, $A_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalised), and then proceeding as usual $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc.
\end{enumerate}
\item this results in the orthonormal basis $A^*$ and a new projection matrix with $P_{m1}=a,~P_{m2}=b$.
\end{enumerate}

@@

\hypertarget{sec:implementation}{%
\section{Implementation}\label{sec:implementation}}

@@

  \begin{itemize}
  \tightlist
  \item
    \texttt{data}: A list of data matrices (they must all have the same
    dimensionality)
  \item
    `\texttt{projmat}: A projection matrix which describes the
    projection plane. This is the initial projection and does not neet
    by orthonormal.
  \item
    `\texttt{xRange}: A list which details the range of the horizontal
    axis.
  \item
    `\texttt{yRange}: A list which details the range of the vertical
    axis.
  \item
    `\texttt{colour}: A list of graphics directives to be applied to
    each data matrix.
  \end{itemize}
\end{itemize}

@@

\textbf{It would be good if the initial projection matrix (and maybe the
centre) had default values. Perhaps a random matrix with the correct
dimensions read from the data?}

@@

\textbf{It would be good if the initial projection matrix (and maybe the
centre) had default values. Perhaps a random matrix with the correct
dimensions read from the data?}

It may be good to fix the size of the points, the centre of the axes and
the scales of the axes rather than allow them to change as the
projection and slicing changes

\begin{itemize}
\item
  \texttt{SliceDynamic{[}data,\ projmat,\ centrePoint,\ height,\ heightRange,\ ptSize1\ :\ 0.005,\ ptSize2\ :\ 0.004{]}}
  with arguments
\item
  \texttt{data}: A data matrix with an extra groupling or clustering
  column (numerical)
\item
  \texttt{projmat}: The inital projection matrix
\item
  \texttt{centrepoint}: The initial position of the slice
\item
  \texttt{height}: The initial height of the slice
\item
  \texttt{heightRange}: The range of heights
\item
  \texttt{ptSize1}: The size of the points that exist within the slice
\item
  \texttt{ptSize2}: The size of the points that exist outside of the
  slice
\end{itemize}

\textbf{Does this need to be a separate function from SlicePlot?}

It would be good if the initial projection matrix (and maybe the centre)
had default values. Perhaps a random matrix with the correct dimensions
read from the data?

It may be good to fix the size of the points, the centre of the axes and
the scales of the axes rather than allow them to change as the
projection and slicing changes

\textbf{I have found it useful to input different data sets rather than
have this function sort them automatically using the last column.}

@@

\textbf{We would need a screenshot (and video) for each of these.}

@@

@@

\begin{itemize}
\tightlist
\item
  As was said before, we need to make sure the drawing range is fixed
  (no resizing of the axis when changing projection or slicing).
\item
  Should also make sure the size of the points is fixed (maybe this will
  be automatic if the drawing range is fixed?), bigger points would be
  preferred, currently the size is sometimes really tiny and it becomes
  difficult to see anything.
\item
  Can we have some heuristic to decide the range for the slice thickness
  slider? In the second graph there is a tiny part of the slider where
  the changes are interesting, difficult to navigate since this makes it
  super sensitive to any small movement, while most of the options are
  not interesting. My suggestion would be to keep (as I understand is
  the case now) the maximum thickness (all points in) as the upper
  bound, but instead of zero use a thickness below which slices tend to
  be empty as the lower bound.
\item
  For the olives example I think the relation between x2 and the blue
  region could be interesting.
\item
  pdfsense data: can we use random samples instead of the first 10k
  rows? Then we should still be able to see the bigger picture (with
  picking the first ones we get some artificial structure from how the
  data was sampled initially).
\end{itemize}

I was playing around with this today. Looks like the axes are fixed when
changing slice thickness, but not when rotating?

Anyways, some observations that might be useful, but all of this seems
difficult for including in a paper:

\begin{itemize}
\tightlist
\item
  In the first example we can show how the yellow group is associated
  with the first variable (x1, I guess this is C9?) - when slicing
  through the center it only appears when x1 has a bigger component on
  the projection, it is associated with low values of x1. Since they are
  not near the average value of x1 the points do not get captured in a
  slice where x1 is not important in the projection. We can move the
  center point to -1 in the first component to then explore how the
  yellow group relates to the other three variables. Much less clear,
  maybe some indication of smaller x2 and larger x3 for this.
\item
  The second example is too busy for manual exploration I would say (too
  many axes to move around separately).
\item
  Third example could be similar to the first: blue region requires
  small values of x2, can be understood from the exploration, then move
  center point to -0.5 in x2 to explore further. Now interpretation
  becomes tricky. Sometimes blue and yellow seem just randomly
  everywhere, sometimes blue more contained in an ellipsoid around the
  center (indicating some correlation between variables). For example
  try zeroing out components 2 and 3 to see this slice in x1 vs x4.
\item
  For the last example x4 seems to be the one to work with. I feel like
  this is again very similar, maybe better to instead explore other
  examples (maybe something without slicing?)
\end{itemize}

>>>>

Manual controls are especially useful for assessing sensitivity of
structure to particular elements of the projection. There are many
places where it is useful. In exploratory data analysis, where one sees
clusters in a projection, can some variables be removed from the
projection without affecting the clustering. For interpreting models,
one can reduce or increase a variable's contribution to examine the
variable importance. These controls can also be used to interactively
generate faceted plots \citep{XXX}, or spatiotemporal glyphmaps
\citep{XXX}. Having the user interact with a projection is extremely
valuable for understanding high-dimensional data. However, these
algorithms have two problems: (1) the pre-processing of creating a
manipulation space overly complicates the algorithm, (2) extending to
higher dimensional control is difficult.

@@

\hypertarget{sec:method}{%
\section{What is a manual tour}\label{sec:method}}

An orthonormal basis (\(A_{p\times d}\)) and a variable id
(\(m \in \{1, ..., p\}\)) to control are provided to initialize a manual
tour. A method to update the values of the component of the controlled
variable \(V_m\) is needed.

\hypertarget{existing-methods}{%
\subsection{Existing methods}\label{existing-methods}}

@@

\hypertarget{the-new-simpler-definition}{%
\subsection{The new simpler
definition}\label{the-new-simpler-definition}}

@@

The problem with Algorithm 1 is that the precise values for \(V_m\)
cannot be specified because the orthonormalisation will change them.
This modification will maintain the components of \(V_m\) precisely
(Figure \ref{fig:manualsequence} (bottom row)). The algorithm is as
follows:

@@

If \(d>2\) the process would be sequentially repeated in the same manner
that Gram-Schmidt is applied sequentially to orthormalise the columns of
a matrix. If \(d=1\) no orthonormalisation is needed, and the projection
vector would simply need to be normalized after each adjustment.

@@

\begin{enumerate}
\item Cursor movement takes ${x_{m1},x_{m2}}\to {a,b}$
\item The freedom to change the components $A_{i3}\cdots A_{ip}$ (the columns of $A$ not corresponding to the projection matrix) is used to select a new orthonormal basis as follows:
\begin{enumerate}
\item For row $m$ one chooses $A_{m3}=\sqrt{1-a^2-b^2},~A_{m,k>3}=0$
\item For other rows, $A_{i\neq m,j\geq3}$ random selections in the range $(-1,1)$ are made.
\item The Gram-Schmidt algorithm is then used to obtain an orthonormal basis taking $e_m$ as the first vector (which is already normalized), and then proceeding as usual $e_1\to e_1-(e_1\cdot e_m) e_m$, $e_1\to e_1/(e_1\cdot e_1)$, etc.
\end{enumerate}
\item this results in the orthonormal basis $A^*$ and a new projection matrix with $P_{m1}=a,~P_{m2}=b$.
\end{enumerate}

@@

\hypertarget{sec:implementation}{%
\section{Experimenting with new techniques using
Mathematica}\label{sec:implementation}}

\begin{itemize}
\tightlist
\item
  \textbf{Why is this a good sandbox}
\item
  \textbf{Explain the functionality available in the notebooks}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    \texttt{data}: A list of data matrices (they must all have the same
    dimensionality)
  \item
    `\texttt{projmat}: A projection matrix which describes the
    projection plane. This is the initial projection and does not need
    to be orthonormal (XXX is this right?).
  \item
    `\texttt{xRange}: A list which details the range of the horizontal
    axis.
  \item
    `\texttt{yRange}: A list which details the range of the vertical
    axis.
  \item
    `\texttt{colour}: A list of graphics directives to be applied to
    each data matrix.
  \end{itemize}
\end{itemize}

@@

\textbf{It would be good if the initial projection matrix (and maybe the
center) had default values. Perhaps a random matrix with the correct
dimensions read from the data? Fix 1}

@@

\textbf{It would be good if the initial projection matrix (and maybe the
center) had default values. Perhaps a random matrix with the correct
dimensions read from the data?}

It may be good to fix the size of the points, the center of the axes and
the scales of the axes rather than allow them to change as the
projection and slicing changes

\begin{itemize}
\item
  \texttt{SliceDynamic{[}data,\ projmat,\ centrePoint,\ height,\ heightRange,\ ptSize1\ :\ 0.005,\ ptSize2\ :\ 0.004{]}}
  with arguments
\item
  \texttt{data}: A data matrix with an extra grouping or clustering
  column (numerical)
\item
  \texttt{projmat}: The initial projection matrix
\item
  \texttt{centrepoint}: The initial position of the slice
\item
  \texttt{height}: The initial height of the slice
\item
  \texttt{heightRange}: The range of heights
\item
  \texttt{ptSize1}: The size of the points that exist within the slice
\item
  \texttt{ptSize2}: The size of the points that exist outside of the
  slice
\end{itemize}

\textbf{Does this need to be a separate function from SlicePlot? Fix 2}

\textbf{It would be good if the initial projection matrix (and maybe the
center) had default values. Perhaps a random matrix with the correct
dimensions read from the data? Fix 3}

\textbf{It may be good to fix the size of the points, the center of the
axes and the scales of the axes rather than allow them to change as the
projection and slicing changes. Fix 4}

\textbf{I have found it useful to input different data sets rather than
have this function sort them automatically using the last column. One
data set with a flag column at the end, is best! Is the code too
complicated? Fix 5}

@@

\textbf{For Alex to do}

\begin{itemize}
\tightlist
\item
  \textbf{We would need a screenshot (and video) for each of these.}
\item
  \textbf{Fix 1, 2, 3, 4, 5}
\item
  \textbf{Code consistent for applications to work - see
  slice\_tour\_compare\_sets.nb. See below}

  \begin{itemize}
  \tightlist
  \item
    \textbf{We need to make sure the drawing range is fixed (no resizing
    of the axis when changing projection or slicing). It could be nice
    if there was a manual zoom in/zoom out to change scale
    occasionally.}
  \item
    \textbf{Should also make sure the size of the points is fixed (maybe
    this will be automatic if the drawing range is fixed?), bigger
    points would be preferred, currently the size is sometimes really
    tiny and it becomes difficult to see anything.}
  \item
    \textbf{Can we have some heuristic to decide the range for the slice
    thickness slider? In the second graph there is a tiny part of the
    slider where the changes are interesting, difficult to navigate
    since this makes it super sensitive to any small movement, while
    most of the options are not interesting. My suggestion would be to
    keep (as I understand is the case now) the maximum thickness (all
    points in) as the upper bound, but instead of zero use a thickness
    below which slices tend to be empty as the lower bound.}
  \item
    \textbf{For the olives example I think the relation between x2 and
    the blue region could be interesting.}
  \item
    \textbf{pdfsense data: can we use random samples instead of the
    first 10k rows? Then we should still be able to see the bigger
    picture (with picking the first ones we get some artificial
    structure from how the data was sampled initially).}
  \item
    \textbf{I was playing around with this today. Looks like the axes
    are fixed when changing slice thickness, but not when rotating?}
  \item
    \textbf{Anyways, some observations that might be useful, but all of
    this seems difficult for including in a paper:}
  \item
    \textbf{In the first example we can show how the yellow group is
    associated with the first variable (x1, I guess this is C9?) - when
    slicing through the center it only appears when x1 has a bigger
    component on the projection, it is associated with low values of x1.
    Since they are not near the average value of x1 the points do not
    get captured in a slice where x1 is not important in the projection.
    We can move the center point to -1 in the first component to then
    explore how the yellow group relates to the other three variables.
    Much less clear, maybe some indication of smaller x2 and larger x3
    for this.}
  \item
    \textbf{The second example is too busy for manual exploration I
    would say (too many axes to move around separately).}
  \item
    \textbf{Third example could be similar to the first: blue region
    requires small values of x2, can be understood from the exploration,
    then move center point to -0.5 in x2 to explore further. Now
    interpretation becomes tricky. Sometimes blue and yellow seem just
    randomly everywhere, sometimes blue more contained in an ellipsoid
    around the center (indicating some correlation between variables).
    For example try zeroing out components 2 and 3 to see this slice in
    x1 vs x4.}
  \item
    \textbf{For the last example x4 seems to be the one to work with. I
    feel like this is again very similar, maybe better to instead
    explore other examples (maybe something without slicing?)}
  \end{itemize}
\end{itemize}

@@

\hypertarget{extension-added-in-the-r-package-tourr}{%
\section{\texorpdfstring{Extension added in the R package
\texttt{tourr}}{Extension added in the R package tourr}}\label{extension-added-in-the-r-package-tourr}}

\hypertarget{what-would-be-desirable-for-implementations-in-r}{%
\section{What would be desirable for implementations in
R?}\label{what-would-be-desirable-for-implementations-in-r}}

@@
